Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2017-11-14 article.Rnw
Text corrections

In practice, overdispersion is largely reported in the literature and
may occur due to the absence of relevant covariates, heterogeneity of
sampling units, different domain size not considered in the analysis and
excess of zeros~\citep{Hinde1998}.  The case of underdispersion is less
report in the literature, however, it has been of increasing interest in
the statistical community. The processes that reduce the variability are
not as well-known as those leading to extra variability.  For the same
reason, there are a few approaches to deal with underdispersed count
data. The explanatory mechanisms leading to underdispersion may be
related to the underline Poisson process generating the count data.
% Here, you gave the idea that you are going to explain all mechanisms.
% Perhaps, you can improve on it!! Have a look at my material for
% regression models school.
When the time between events is not exponential distributed, the number
of events can be over or underdispersed, this process motivated the
class of duration dependence models~\citep{Winkelmann1995}.  Another
possibility to explain the underdispersion phenomenon is order
statistics.  Suppose $Y_i$ is Poisson distributed then the maximum of
$Y_i$ is underdispersed. (NEED REF).

@@

\footnotesize \raggedright
divergent$^{*}$ is a mathematically divergent serie; and
divergent$^{**}$ a numerically divergent serie.
\end{table}

In the first line of \autoref{tab:convergenceZ} we have mathematically
divergent series, because $\sum_{j=0}\lambda^y$, when $\lambda \geq 1$,
is divergent.  In other cases the series diverges numerically, due to
the computational storage limitation.  For both divergences it is
impossible to compute probabilities, therefore, this acts as a
restriction in the parameter space.

An undesirable feature of the COM-Poisson distribution is that the
moments cannot be obtained in closed form. \citet{Shmueli2005} and
\citet{Sellers2010} using an asymptotic approximation for
$Z(\lambda,\nu)$, showed that the expectation and variance of the
COM-Poisson distribution can be approximated by
\begin{equation}
  \label{eqn:mean-approx}
  E(Y) \approx \lambda^{1/\nu} - \frac{\nu - 1}{2\nu} \qquad
  \textrm{and} \qquad
  \textrm{var}(Y) \approx \frac{\lambda^{1/\nu}}{\nu}\,,
\end{equation} which is particularly accurate for  $\nu \leq 1$ or
$\lambda > 10$. The authors also argue that the mean-variance
relationship can be approximate by
$\frac{1}{\nu}E(Y)$. In~\autoref{reparametrization} we assess the
accucary of these approximations.

@@

These indexes are discussed by \citet{Bonat2017} to study the
flexibility of Poisson-Tweedie distribution, and \citet{Puig2006} to
describe count distributions. Regarding the COM-Poisson$_\mu$
distribution, we presents in \autoref{fig:indexes-plot} the relationship
between mean and variance (a), the dispersion and zero-inflation indexes
(b--c) for differents values of $\mu$ and $\phi$ and heavy-tail index
(d) for $\mu=25$ and differents values of $y$ and $\phi$.

@@

\autoref{fig:indexes-plot} shows that the indexes are slightly depend on
the expected values and tend to stabilize for large values of
$\mu$. Consequently, the mean and variance relationship
\autoref{fig:indexes-plot}(a) is proportional to the dispersion
parameter $\phi$. In terms of moments, this leads to a specification
indistinguishable from the quasi-Poisson regression model. The
dispersion indexes in \autoref{fig:indexes-plot}(b) show that the
distribuion is suitable to deal to dispersed counts, of course. For the
parameter values evaluated the largest DI was \Sexpr{madi} and smallest
was \Sexpr{midi}. \autoref{fig:indexes-plot}(c) shows the COM-Poisson
can handle with a limited amount of zero-inflation, in cases of
overdispersion ($\phi < 0$). On the other hand, for $\phi > 0$
(underdispersion) this distribution is suitable to deal with
zero-deflated counts. Heavy-tail indexes in
\autoref{fig:indexes-plot}(d) indicates the distribution is in general
light tailed distribution, i.e. $HT \to 0$ for $y \to \infty$.

@@

The observed and fitted counts for each humidity level with confidence
intervals are shown in \autoref{fig:pred-soy}. The fitted values are
identical for the Poisson and COM-Poisson$_{\mu}$ models, leading to the
same conclusions. On the other hand, confidence intervals for the
Poisson model are smaller than the ones from the
COM-Poisson$_{\mu}$. Probably, it is due to the equidispersion
assumption underline the Poisson model. The confidence intervals from
the COM-Poisson$_{\mu}$ and quasi-Poisson models are really similar,
which in turn show the already highlighted similarity between these
approches.

@@

The data set comes from an experiment to measure the reproductive
toxicity of a herbicide, nitrofen, on a species of zooplankton
(\textit{Ceriodaphnia dubia}). Fifty animals were randomized into
batches of ten and each batch was put in a solution with a measured
concentration of nitrofen (0, 0.8, 1.6, 2.35 e 3.10 mug$/10^2$litre)
(\texttt{dose}). Then the number of total live offspring was recorded
\cite{Bailer1994}.

@@

For this data set we proposed three linear predictors. We compare the
performance of models using the likelihood ratio tests. The linear
predictors are
\begin{center}
\begin{minipage}{12cm}
Predictor 1: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i$\\
Predictor 2: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2$\\
Predictor 3: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3$.
\end{minipage}
\end{center}

@@

In general, the results presented by the reparametrized COM-Poisson
models were satisfactory and comparable to the convential approaches.
Therefore, its use in the analysis of count data is encouraged. The
computational routines for fitting the original and reparametrized
COM-Poisson regression models are available in the supplementary
material\textsuperscript{\ref{papercompanion}}.

>>>>

In practice, overdispersion is largely reported in the literature and
may occur due to the absence of relevant covariates, heterogeneity of
sampling units, different domain size not considered in the analysis and
excess of zeros~\citep{Hinde1998}.  The case of underdispersion is less
report in the literature, however, it has been of increasing interest in
the statistical community. The processes that reduce the variability are
not as well-known as those leading to extra variability.  For the same
reason, there are a few approaches to deal with underdispersed count
data. The explanatory mechanisms leading to underdispersion may be
related to the underline Poisson process generating the count data.
When the time between events is not exponential distributed, the number
of events can be over or underdispersed, this process motivated the
class of duration dependence models~\citep{Winkelmann1995}.  Another
possibility to explain the underdispersion phenomenon is order
statistics.  Suppose $Y_i$ is Poisson distributed then the maximum of
$Y_i$ is underdispersed. (NEED REF).

@@

\footnotesize \raggedright
divergent$^{*}$ is a mathematically divergent series; and
divergent$^{**}$ a numerically divergent aerie.
\end{table}

In the first line of \autoref{tab:convergenceZ} we have mathematically
divergent series, because $\sum_{j=0}^\infty\lambda^y$, when
$\lambda \geq 1$, is divergent.  In other cases the series diverges
numerically, due to the computational storage limitation.  For both
divergences it is impossible to compute probabilities, therefore, this
acts as a restriction in the parameter space.

An undesirable feature of the COM-Poisson distribution is that the
moments cannot be obtained in closed form. \citet{Shmueli2005} and
\citet{Sellers2010} using an asymptotic approximation for
$Z(\lambda,\nu)$, showed that the expectation and variance of the
COM-Poisson distribution can be approximated by
\begin{equation}
  \label{eqn:mean-approx}
  E(Y) \approx \lambda^{1/\nu} - \frac{\nu - 1}{2\nu} \qquad
  \textrm{and} \qquad
  \textrm{var}(Y) \approx \frac{\lambda^{1/\nu}}{\nu}\,,
\end{equation} which is particularly accurate for  $\nu \leq 1$ or
$\lambda > 10$. The authors also argue that the mean-variance
relationship can be approximate by
$\frac{1}{\nu}E(Y)$. In~\autoref{reparametrization} we assess the
accuracy of these approximations.

@@

These indexes are discussed by \citet{Bonat2017} to study the
flexibility of Poisson-Tweedie distribution, and \citet{Puig2006} to
describe count distributions. Regarding the COM-Poisson$_\mu$
distribution, we presents in \autoref{fig:indexes-plot} the relationship
between mean and variance (a), the dispersion and zero-inflation indexes
(b--c) for different values of $\mu$ and $\phi$ and heavy-tail index
(d) for $\mu=25$ and different values of $y$ and $\phi$.

@@

\autoref{fig:indexes-plot} shows that the indexes are slightly depend on
the expected values and tend to stabilize for large values of
$\mu$. Consequently, the mean and variance relationship
\autoref{fig:indexes-plot}(a) is proportional to the dispersion
parameter $\phi$. In terms of moments, this leads to a specification
indistinguishable from the quasi-Poisson regression model. The
dispersion indexes in \autoref{fig:indexes-plot}(b) show that the
distribution is suitable to deal to dispersed counts, of course. For the
parameter values evaluated the largest DI was \Sexpr{madi} and smallest
was \Sexpr{midi}. \autoref{fig:indexes-plot}(c) shows the COM-Poisson
can handle with a limited amount of zero-inflation, in cases of
overdispersion ($\phi < 0$). On the other hand, for $\phi > 0$
(underdispersion) this distribution is suitable to deal with
zero-deflated counts. Heavy-tail indexes in
\autoref{fig:indexes-plot}(d) indicates the distribution is in general
light tailed distribution, i.e. $HT \to 0$ for $y \to \infty$.

@@

The observed and fitted counts for each humidity level with confidence
intervals are shown in \autoref{fig:pred-soy}. The fitted values are
identical for the Poisson and COM-Poisson$_{\mu}$ models, leading to the
same conclusions. On the other hand, confidence intervals for the
Poisson model are smaller than the ones from the
COM-Poisson$_{\mu}$. Probably, it is due to the equidispersion
assumption underline the Poisson model. The confidence intervals from
the COM-Poisson$_{\mu}$ and quasi-Poisson models are really similar,
which in turn show the already highlighted similarity between these
approaches.

@@

The data set comes from an experiment to measure the reproductive
toxicity of a herbicide, nitrofen, on a species of zooplankton
(\textit{Ceriodaphnia dubia}). Fifty animals were randomized into
batches of ten and each batch was put in a solution with a measured
concentration of nitrofen ($0, 0.8, 1.6, 2.35$ e $3.10$ mug$/10^2$litre)
(\texttt{dose}). Then the number of total live offspring was recorded
\cite{Bailer1994}.

@@

For this data set we proposed three linear predictors. We compared the
performance of models using the likelihood ratio tests. The linear
predictors are
\begin{center}
\begin{minipage}{12cm}
Predictor 1: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i$\\
Predictor 2: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2$\\
Predictor 3: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3$.
\end{minipage}
\end{center}

@@

In general, the results presented by the reparametrized COM-Poisson
models were satisfactory and comparable to the conventional approaches.
Therefore, its use in the analysis of count data is encouraged. The
computational routines for fitting the original and reparametrized
COM-Poisson regression models are available in the supplementary
material\textsuperscript{\ref{papercompanion}}.
